Lossy compression of DNA sequencing quality data

### 2015-05-11 manuscript/q-compression.tex
Lots of polishing of the text. We are very close to a submittable version with this draft.

The \textsc{fastq} file format has become the \emph{de facto} standard
for storing next-generation sequencing data, containing nucleotide
information along with a quantitative measure of the reliability of
individual base calls. As the cost of sequencing continues to
decrease, the rate of sequence data production is increasing,
requiring efficient ways of storing and transferring this vast amount
of data. Most methods of sequencing data compression focus on
compressing nucleotide information without any loss of information.
Quality data, however, have different properties than nucleotide data,
and methods compressing nucleotide sequences efficiently do not
perform as well on quality sequences. Furthermore, while lossless
representation is necessary for nucleotide sequences, it is not an
essential requirement for quality values.

@@

Quality values can be used throughout bioinformatics pipelines. Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, which
removes reads that on the whole do not meet arbitrary quality
standards, thus reducing the total number of reads; and
\textit{ii}.~trimming of low-quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}. Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies. Sequence reads generated by instruments such
as an Illumina HiSeq, the focus of this research due to its
ubiquitousness in bioinformatics research, are characterized by having
relatively few insertion and deletion errors, but substitution
(miscall) errors are much more frequent and have context-specific
patterns. These errors are non-uniformly distributed over the read
length (e.g., error rates increase up to $\sim$16$\times$ at the
3$^{\prime}$ end, and 32.8 -- 67.9\% of reads have low-quality tails
at the 3$^{\prime}$ end~\citep{Minoche:2011km}).

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective, some loss of information
is deemed acceptable if the inferences from analyses are relatively
unaffected. Here we describe our research investigating lossy
compression of sequence read quality values, specifically those
associated with Illumina instruments, with the objective to provide
some perspective on several strategies rather than to develop robust
high-quality software for use. Recognizing these properties of
Illumina sequence reads motivates our exploration of three general
classes of lossy compression methods -- binning, modeling, and
profiling -- and our consideration of an exemplar of each
class.~\cite{Canovas:2014fr} and~\cite{janin2013adaptive} evaluated
the effects of lossy compression on identifying variants within a data
set. We build on these prior works, and assess the effects of quality
value information loss resulting from compression on subsequent
genomic analyses, including read preprocessing (filtering and
trimming), genome assembly, and read mapping.

@@

\cite{Wan:2012kq} provide three similar lossy compression strategies
based on binning the base error probability distribution: UniBinning,
Truncating, and LogBinning. UniBinning evenly splits the error
probability distribution into a user-defined number of partitions.
Truncating treats a user-defined number of highest quality values as a
single bin. LogBinning works similarly to UniBinning, except it uses
the \emph{log} of the error probability distribution, which
effectively bins the \textsc{ascii} quality values evenly. Our 2-bin
encoding is a combination of LogBinning and Truncating in that we
place the highest quality values (as defined above) of the log of the
error probability distribution into a single bin.

@@

While we focus only on two bins in this work, a greater number of bins
can be used in practice. For example, an additional third bin (good,
bad, and \emph{okay}) can be used for base qualities near the border
between bins in the 2-bin encoding. Examining the trade-off of
additional bins and their resulting compressibility and effect on
downstream analyses is outside the scope of this work.

@@

Sets of quality strings show similar trends of quality over their
length, and it is possible to identify common patterns in the data and
use them as reference profiles to approximate individual sequences of
quality values. Here we use $k$-means clustering, a vector
quantization method that partitions a set of samples into $k$ sets
that minimize within-cluster sum of
squares~\citep{macqueen1967some}. We sampled 10$K$ reads at random and
computed cluster centers as read quality profiles using a heuristic
iterative refinement approach that quickly converges to a locally
optimal minimum~\citep{hartigan1979algorithm}.  All reads are
evaluated with the nearest quality profile in Euclidean space being
assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read-specific differences in quality values, it
does approximate the overall patterns in quality values. An example of
128 quality profiles is shown in Figure \ref{fig:profiles_128}.

The economic costs of memory use for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is a function of the number of profiles times the cost
of encoding them, and these fixed costs are amortized over the entire
set of reads to which they apply. Additionally there are marginal
costs that are a function of the number of reads encoded.

Profiles can be obtained from techniques other than $k$-means
clustering. We can store the profiles of polynomial functions to
increase the compressibility of polynomial regression modeling.  In
other words, we can use a spline (a function that is piecewise-defined
by polynomial functions) to represent a given quality sequence.
However, we have not explored any of these other approaches here.

@@

We used several Illumina sequence read data sets in this research,
which are taken from data from the \textsc{gage} (Genome Assembly
Gold-Standard Evaluations)~\citep{Salzberg:2012rc}, except as
noted. These data sets are as follows.

@@

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180bp nt;
1,145,8940 paired-end reads). The corresponding reference sequence was
obtained from the \textsc{ncbi} RefSeq database (NC\_000913.2).

@@

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool~\citep{Ochoa:2013rt}. The program models
quality values as a multivariate Gaussian distribution, computing the
mean and covariance for each position in the read, and these varlues
are stored by the decoder to later reconstruct a \emph{representative}
quality value. \textsc{QualComp} takes as input a user-specified rate
(bits/read), and these bits are allotted among positions within the
read to minimize the average error. The quality values can be
clustered beforehand to produce more accurate models.

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window ($0.1 \times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the
respective end. If the trimmed sequence length is less than a certain
threshold (20 by default), then the sequence is discarded.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include N50, which
is defined as the median contig size (the length of largest contig $c$
such that the total size of the contigs larger than $c$ exceeds half
of the assembly size) and corrected N50, which is the recalculated N50
size after the contigs are broken apart at locations of errors. The
\textsc{lap} score can be viewed as a log-likelihood score, where a
value closer to 0 is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use of quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

We compare the \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets (Figure
\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets,
but the short-jump library results are available in Supplemental Table
1. Storing the uncompressed quality values requires 1 byte/base-call
because they are stored in \textsc{ascii} format, which is denoted by
the dotted black asterisk in the figure. The lossless compression of
each data set using \textsc{bz}ip2 ranges from 2.19--3.10
bits/base-call and is denoted by the colored asterisks on the
abscissa. The compression methods tend to cluster together across the
different data sets. Across all data sets, the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more storage than the profile encoding methods
for only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10
bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas
256-profile encoding only requires 0.079 bits/bp and has a
\textsc{mse} of 11.85.

@@

Preprocessing involves two steps: \emph{trimming} the poor-quality
regions of the sequences, and \emph{discarding} sequences that are
deemed to be poor-quality overall. After trimming, the majority of
compression methods retain more base-pairs than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression model increases in complexity --- i.e., as the number of
profiles, polynomial degrees, or rate increases --- the amount of
base-pairs kept approaches the number of base-pairs kept using the
uncompressed sequences. The compression methods on the \textit{Mus
  musculus} data set have the greatest proportion of retained
base-pairs compared to the uncompressed sequences. The
\textit{Escherichia coli} MiSeq data set has the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-pairs across all data sets. Sickle uses
a sliding window approach to smooth the read quality values before it
trims. In the 2-bin approach, there is not an even distribution of
values per bin. In other words, \emph{bad} quality values may range
from 0--33, whereas \emph{good} values may only range from
34--40. Thus, mid-range quality values that are above the threshold
(20 by default) are set below the quality threshold when compressed,
resulting in an increased number of trimmed bases.

The 0-degree polynomial regression results in the highest proportion
of bases kept. If the mean quality value of the read is above the
filtering threshold, then no base-pairs are trimmed. Only reads that
are comprised of mainly low quality values will be discarded.

It is important to highlight that even though a compression method may
result in the same number of trimmed bases as the uncompressed
sequences, it does not mean the \emph{same} bases were trimmed. The
1st-degree and 5th-degree polynomial regression of the
\textit{Rhodobacter sphaeroides} fragment library filter nearly as
many bases as each other. However, if we examine the specific reads
discarded, the 5th-degree polynomial regression discards approximately
two-thirds less reads than the 1st-degree polynomial regression that
are kept by the uncompressed reads (4,640 and 12,066 reads,
respectively) (Supplemental Table 2).

@@

No assembly of the \textit{Rhodobacter sphaeroides} data set was best
in all metrics (Table \ref{fig:assembly_ranks}). Among the compression
methods, the profile encoding performed best, polynomial regression
modeling performed worst, and other methods were intermediate
(Fig.~\ref{fig:assembly_ranks}).

@@

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads score worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using the original uncompressed reads has an error rate of
roughly 8.75 errors per 100 kb of assembled sequence, while the
256-profile encoding has an error rate of 8.02 errors per 100 kbp
(Supplemental Table 3).

@@

Ranking the compression methods by overall alignment rate produces an
identical ordering as above. Aside from 0-degree polynomial regression
(83.1\%), all other compression methods have an alignment rate between
87\% and 86.1\%. The alignment rate of the uncompressed reads is 87\%.

@@

Setting all bases as minimum quality results in the highest proportion
of mapped reads (88.1\%). Conversely, setting all bases as maximum
quality results in the lowest proportion of mapped reads (72.8\%).

@@

It is common to evaluate lossy compression methods in terms of
information loss, and some measure of compression effectiveness.
However, from a pragmatic perspective, it is the effects on the
results of subsequent analyses that are most important. Our results
showed that some of the compression methods with relatively high
information lossy (\textsc{mse}) are still practical for some
applications. Many bioinformatics tools proved to be robust to
moderate information loss. Passing the decompressed quality values
through quality control software shows that most methods filter nearly
as many bases as using original quality sequences. Assemblers
performing sequence alignment use percent similarity scores that are
typically robust to standard sequencing errors.

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low-quality bases that occur within stretches of
high-quality bases, whereas the polynomial regression methods tend to
smooth these qualities. In addition to aligning fewer reads than the
2-bin encoding, the polynomial regression methods may also be
hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality may lead to the base being incorrectly labeled as a
\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics.

@@

In this paper we have examined several general approaches for lossy
compression of sequence quality values and their effect on subsequent
analyses. Although most previous examinations of lossy compression
focused primarily on information loss and compression effectiveness,
we show that common bioinformatic analyses are robust to substantial
loss of information in quality values.

@@

\section*{Acknowledgement}
This research was initiated at the 2014 Bioinformatics Exchange for
Students and Teachers (\textsc{best}) Summer School, which was funded
by the offices of the Dean of The Graduate School at University of
Maryland and the Rektor of University of T\"{u}bingen.

>>>>

As the cost of sequencing continues to decrease, the rate of sequence
data production is increasing, placing greater demands on storing and
transferring these vast amount of data. Most methods of sequencing
data compression focus on compressing nucleotide information without
any loss of information.  Quality data, however, have different
properties than nucleotide data, and methods compressing nucleotide
sequences efficiently do not perform as well on quality
sequences. Furthermore, while lossless representation is necessary for
nucleotide sequences, it is not an essential requirement for quality
values.

@@

Quality values can be used throughout bioinformatics pipelines. Among
the most fundamental uses of sequence quality values is as part of the
quality assessment and quality control (\textsc{qa/qc}) processes
prior to subsequent analysis steps. Quality control based on quality
values generally includes two operations: \textit{i}.~filtering, which
removes reads that on the whole do not meet arbitrary quality
standards, thus reducing the total number of reads; and
\textit{ii}.~trimming of low-quality base-calls from reads, which
reduces the length of reads trimmed. Quality values can be used by
genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment programs, such as Bowtie2~\citep{Langmead:2012rw}, use
quality values to weight mismatches between read and reference
sequences. Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values~\cite[e.g.,][]{McKenna:2010bh},
and identified \textsc{snp}s with high-quality calls are deemed more
reliable than those with low-quality calls, particularly in
low-coverage regions.

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}. Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective, some loss of information
is deemed acceptable if the inferences from analyses are relatively
unaffected. Sequence reads generated by instruments such as an
Illumina HiSeq, the focus of this research due to its preponderance in
genomics research, are characterized by having relatively few
insertion and deletion errors, but substitution (miscall) errors are
much more frequent and have context-specific patterns. These errors
are non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
and~\cite{janin2013adaptive} evaluated the effects of lossy
compression on identifying variants within a data set. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We build on these prior works, and assess the effects of quality
value information loss resulting from compression on subsequent
genomic analyses, including read preprocessing (filtering and
trimming), genome assembly, and read mapping.

@@

\cite{Wan:2012kq} provide three similar lossy compression strategies
based on binning the base error probability distribution: UniBinning,
Truncating, and LogBinning. UniBinning evenly splits the error
probability distribution into a user-defined number of partitions.
Truncating treats a user-defined number of highest quality values as a
single bin. LogBinning works similarly to UniBinning, except it uses
the \emph{log} of the error probability distribution, which
effectively bins the \textsc{ascii} quality values evenly. The 2-bin
encoding examined here can be viewed as something like a combination
of LogBinning and Truncating in that we place the highest quality
values (as defined above) of the log of the error probability
distribution into a single bin.

@@

While we focus only on two bins in this work, a greater number of bins
can be used in practice. For example, an additional third bin (good,
bad, and intermediate) can be used for base qualities near the border
between bins in the 2-bin encoding. Examining the trade-off of
additional bins and their resulting compressibility and effect on
downstream analyses are not considered here.

@@

Sets of strings representing quality values show similar trends of
over their length, and it is possible to identify common patterns in
the data and use them as reference profiles to approximate individual
sequences of quality values. Here we use $k$-means clustering, a
vector quantization method that partitions a set of samples into $k$
sets that minimize within-cluster sum of
squares~\citep{macqueen1967some}. We sampled 1 $\times$ 10$^{4}$ reads
at random and computed cluster centers as read quality profiles using
a heuristic iterative refinement approach that quickly converges to a
locally optimal minimum~\citep{hartigan1979algorithm}. All reads are
evaluated and the nearest quality profile in Euclidean space being
assigned to every read as their compressed representation.  The
compressed quality file therefore consists of an index enumerating the
$k$ quality profiles, and a binary part containing the assigned
quality profile index for each read.  Although this approach is not
able to capture some read-specific differences in quality values, it
does approximate the overall patterns in quality values. An example of
128 quality profiles is shown in Figure \ref{fig:profiles_128}.

The economic costs of memory use for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is a function of the number of profiles times the cost
of encoding them, and these fixed costs are amortized over the entire
set of reads to which they apply. Additionally there are marginal
costs that are a function of the number of reads encoded times the
cost of the encoding.  Profiling like modeling, but not binning, has
the advantageous property that cost improves with read length.

Profiles can be obtained from techniques other than $k$-means
clustering. For example, we could store the profiles of polynomial
functions to increase the compressibility of polynomial regression
modeling.  In other words, we could use a spline (a function that is
piecewise-defined by polynomial functions) to represent a given
quality sequence.  However, we have not explored these other
approaches here.

@@

We used several Illumina sequence read data sets taken from data from
the \textsc{gage} (Genome Assembly Gold-Standard
Evaluations)~\citep{Salzberg:2012rc}, except as noted. These data sets
are as follows.

@@

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180 nt;
1,145,8940 paired-end reads). The corresponding reference sequence was
obtained from the \textsc{ncbi} RefSeq database (NC\_000913.2).

@@

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool~\citep{Ochoa:2013rt}. The program models
quality values as a multivariate Gaussian distribution, computing the
mean and covariance for each position in the read, and these varlues
are stored by the decoder to later reconstruct a \emph{representative}
quality value. \textsc{QualComp} takes as input a user-specified rate
(bits/read), and these bits are apportioned among positions within the
read to minimize the average error. The quality values can be
clustered beforehand to produce more accurate models.

@@

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window ($0.1 \times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the read
end. If the trimmed sequence length is less than a certain threshold
(20 by default), then the sequence is discarded.

We evaluate effects of information loss from quality value compression
on \emph{de novo} genome assembly performance using contiguity
statistics, log average read probability
(\textsc{lap})~\citep{Ghodsi:2013hb}, and a collection of
reference-based metrics. The contiguity statistics include N50, which
is defined as the median contig size (the length of largest contig $c$
such that the total size of the contigs larger than $c$ exceeds half
of the assembly size) and corrected N50, which is the recalculated N50
size after the contigs are broken apart at locations of errors. The
\textsc{lap} score can be viewed as a log-likelihood score, where a
larger value is better. We use a script provided by \textsc{gage}
reference-based evaluation to count single nucleotide polymorphisms
(\textsc{snp}s), relocations, translations, and inversions. The
reference-based metrics are normalized by the length of the assembly
to facilitate comparison. For the genome assembly we used software
that makes use of quality values in the assembly process:
\textsc{allpaths-lg}~\citep{Gnerre:2011kx} version r50191 with default
settings and 32 threads.

@@

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations, which include both fixed and
marginal costs (Figure \ref{fig:mse_vs_bpbp}). Here we present the
fragment libraries for the \textit{Rhodobacter sphaeroides}, and
\textit{Homo sapiens} data sets; the short-jump library results are
available in Supplemental Table 1. Storing the uncompressed quality
values requires 1 byte/base-call because they are stored in
\textsc{ascii} format, which is denoted by the dotted black asterisk
in the figure. The lossless compression of each data set using
\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call and is denoted by
the colored asterisks on the abscissa; these values provide a
reference point using a widely available compression program. Values
from the same class of compression method tend to cluster together
across the different data sets, with the 0-degree polynomial
regression, profile encodings, and \textsc{q}ual\textsc{c}omp have the
lowest bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more memory than the profile encoding methods for
only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10
bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas
256-profile encoding only requires 0.079 bits/bp and has a
\textsc{mse} of 11.85.

@@

Preprocessing involves two steps: \emph{discarding} reads that are
deemed to be poor-quality overall, and \emph{trimming} the
poor-quality regions of the reads. After trimming, the majority of
compression methods retain more base-calls than the uncompressed
sequences (Figure \ref{fig:preprocessing}). In general, as a given
compression method increases in complexity --- i.e., as the number of
profiles, polynomial degrees, or rate increases --- the amount of
base-calls retained more closely approximates the number of base-calls
retained using the uncompressed sequences. The compression methods on
the \textit{Mus musculus} data set had the greatest proportion of
retained base-calls compared to the uncompressed sequences. The
\textit{Escherichia coli} MiSeq data set had the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-calls compared to the original
uncompressed reads across all data sets. Sickle uses a sliding window
approach to smooth the read quality values before it trims. In the
2-bin approach, there is an uneven distribution of values per bin. In
other words, \emph{bad} quality values may range from 0--33, whereas
\emph{good} values may only range from 34--40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
trimmed bases.

The 0-degree polynomial regression results in the highest proportion
of base-calls kept. If the mean quality value of the read is above the
filtering threshold, then no base-calls are trimmed. Only reads that
are comprised of mostly low quality values will be discarded.

It is important to highlight that even though a compression method may
result in the same number of trimmed base-calls as the uncompressed
sequences, it does not mean the \emph{same} base-calls were retained.
For example, pre-processing of the 1st-degree and 5th-degree
polynomial regression model compressed reads of the
\textit{Rhodobacter sphaeroides} fragment library retain approximately
the same number of base-calls. However, if we examine the specific
reads discarded, the 5th-degree model discards approximately
two-thirds less reads than the 1st-degree model (4,640 and 12,066
reads, respectively) (Supplemental Table 2).

@@

No compression method resulted in the uniformly best assembly of the
\textit{Rhodobacter sphaeroides} data set in all metrics (Table
\ref{fig:assembly_ranks}). Among the compression methods, the profile
encoding performed best, polynomial regression modeling performed
worst, and other methods were intermediate
(Fig.~\ref{fig:assembly_ranks}).

@@

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads score worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using the original uncompressed reads has an error rate of
$\sim$8.75 errors/100 kb of assembled sequence, while the 256-profile
encoding has an error rate of $\sim$8.02 errors/100 kbp (Supplemental
Table 3).

@@

Ranking the compression methods by overall proportion of reads aligned
produces an identical ordering as above. Aside from 0-degree
polynomial regression (0.831), all other compression methods have a
read alignment rate between 0.87 and 0.861. The proportion of reads
aligned for the uncompressed reads is 0.87.

@@

Setting all bases as minimum quality results in the highest proportion
of mapped reads (0.881). Conversely, setting all bases as maximum
quality results in the lowest proportion of mapped reads (0.728).

@@

It is common to evaluate lossy compression methods in terms of
information loss, and some measure of compression effectiveness.
However, from a pragmatic perspective, it is the consequences for
subsequent analyses that are most important. Our results showed that
some of the compression methods with relatively high information lossy
(\textsc{mse}) are practical, and possibly advantageous, for some
applications. Many bioinformatics tools proved to be robust to
moderate information loss. Passing the decompressed quality values
through quality control software shows that most methods filter nearly
as many bases as using original quality sequences. Assembly programs
performing sequence alignment use percent similarity scores that are
typically robust to standard sequencing errors.

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low-quality bases that occur within stretches of
high-quality bases, whereas the polynomial regression methods tend to
smooth these qualities. In addition to aligning fewer reads than the
2-bin encoding, the polynomial regression methods may also be
hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality may lead to the base being incorrectly labeled as a
\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics. Profiling
performs quite well in all use cases.

@@

We have examined several simple and general approaches for lossy
compression of sequence quality values and their effect on subsequent
analyses. Although most previous examinations of lossy compression
focused primarily on information loss and compression effectiveness,
we show that common bioinformatic analyses are robust to substantial
loss of information in quality values. Some methods, such as
profiling, have yield excellent compression and provide excellent
results in subsequent analyses, and in some cases surpassing results
from original uncompressed data.

@@

\section*{Acknowledgement}
This project was initiated at the 2014 Bioinformatics Exchange for
Students and Teachers (\textsc{best}) Summer School, which was funded
by the offices of the Dean of The Graduate School, University of
Maryland, and the Rektor of the University of T\"{u}bingen. Additional
funding included an International Graduate Research Fellowship from
The Graduate School, University of Maryland, to CMH, a Global
Partnerships-Faculty Travel Grant from the Office of International
Affairs to MPC, and other funding from the University of T\"{u}bingen.
